---
title: "NYPD Shooting Data Report"
author: "John Mitchell"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results="hide",warning=FALSE, message=FALSE)

# Load libraries needed:
library(tidyverse)
```

### Required R Packages 
The `tidyverse` library is loaded at the start of the document. If the package is not installed, use `install.packages()` to install it before knitting.

# 1. Overview

This report explores the New York Police Department (NYPD) data set of shootings from 2006 to 2023, and focuses on how average incident rates differ across the five different boros (or "boroughs") of New York, that is: Staten Island, Queens, Manhattan, the Bronx, and Brooklyn. The questions of interest are:

1. Did the rate of shooting incident - adjusted for boro population - differ across the five boros over the time period? 

2. Looking the incident rates in each boro over time, are there any noticeable patterns?

To investigate these issues:

* Section 2 imports, tidies, and summarizes the data.
* Section 3 explores average shooting incident rates across the boros visualisations, and more formally through a one-way ANOVA (Analysis of Variance).
* Section 4 turns to the second question - trends over time, and looks for potential patterns using time series plots.
* Then, section 5 discusses potential sources of bias, both personal and in in the data source.
* Section 6 Summarizes the conclusions.

Two additional sections - appendices A and B - go into more detail on: how the data set handles incidents with multiple victims, and how the boro population data (which is not part of the raw data set) were obtained.

References are directly linked to when first mentioned.

**R libraries needed** 

# 2. Data Importing, Tidying, and Summarizing

The following sections import the data, summarize overall features, and tidy it for later analysis.

## 2.1 Importing Raw Data

The data set is imported below.  Each row of the data set is a record of a shooting incident in one of the five boros (administrative districts) of New York City from January 2006 to December 2023.


The code below loads the raw data from the repository at "https://opendata.cityofnewyork.us/". Output is not shown.

```{r load_nypd}

raw_data <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")

ny_initial_processing <- raw_data %>%
    select(c("INCIDENT_KEY", "OCCUR_DATE", "OCCUR_TIME", "BORO",
             "STATISTICAL_MURDER_FLAG","VIC_AGE_GROUP", "VIC_SEX", "VIC_RACE")) %>%
    rename(`key`=`INCIDENT_KEY`,`flag`=`STATISTICAL_MURDER_FLAG`,'boro'='BORO') %>%
    mutate(date=as.Date(OCCUR_DATE,format="%m/%d/%Y" )) %>%
    select(!c("OCCUR_DATE", "OCCUR_TIME"))

# change one typo in an age to unknown
ny_initial_processing[ny_initial_processing$VIC_AGE_GROUP==1022,]$VIC_AGE_GROUP="UNKNOWN"

# Convert the "victim" columns to factors
ny_victim_factors <- c("VIC_AGE_GROUP", "VIC_SEX", "VIC_RACE")
ny_initial_processing[ny_victim_factors] <- lapply(ny_initial_processing[ny_victim_factors],factor)


```

# 2.2 Summary of Raw Data

The analysis will focus on a subset of the original data. However the structure and content of the original raw data set will be outlined briefly. 

* The original data set has 28562 records of shooting incidents over the time period January 2006 to December 2023. 
* There are 21 columns in the data set, so each record can have up to 21 attributes.
* A data dictionary same source as the data set describes the columns in detail.

The columns include:

* An unique incident key for each incident. More details about this key - and its limitations - are discussed below. 
* The Date and time of the incident
* What boro and precinct (police adminstrative district) the incident took place in.
* Several columns with short verbal descriptions of the location (if known) such as "outside". In most cases these fields are missing.
* Detailed coordinate information (latitude and longitude). The data dictionary describes in detail how this is calculated.
* Demographic information (such as gender, age group and race) of the victim, if known.
* Demographic information (such as gender, age group and race) of the perpetrator of the shooting, if known.
* A true or false "flag" of whether the incident was a murder - that is, if there has been a death in the shooting incident.

The incident key is unique if there was one victim of the shooting. For incidents with multiple victims, multiple rows are created with the same incident key. The data dictionary states that "A shooting incident can have multiple victims involved and as a result duplicate INCIDENT_KEY’s are produced. Each INCIDENT_KEY represents a victim but similar duplicate keys are counted as one incident". However, the lack of a secondary key uniquely identifying each victim made quickly tallying mass shooting incidents difficult. Brief notes for those interested are in the "Additional Notes on Incidents with Multiple Victims" section at the end of the document.

It is also worth noting information *not* present in the data set. Such items include:

* Any context that would help determine *why* the shooting took place.
* Any information about the severity of the shooting (beyond whether there was a death at the scene).
* For any demographic information about the perpetrator (or supposed perpetrator), any indication of how that information is known and its reliability.
* Any contact information to check individual records for accuracy.


# 2.3 Including Boro Population Data

As part of the analysis the shootings in each boro will be scaled by the population of that boro. Estimates of boro populations were obtained from the 2020 census. See Appendix A for details on the raw data source. The data is in the table below.


```{r boro_data, echo=FALSE, results="markup"}

boro_populations = data.frame(boro=c("MANHATTAN","BRONX","BROOKLYN","QUEENS","STATEN ISLAND"), population = c(1694251,1472654,2736074,2405464,495747))

knitr::kable(boro_populations) %>% kableExtra::kable_styling(position = "center",full_width=F) 

```



# 2.4 Data Tidying 

In the code below, a subset of the data is extracted for detailed analysis. 

* A data frame of unique shooting events with the time of each, and the boro it occurred in, is created. The analysis will focus on shooting events in each boro - irrespective of the number of casualties. 
* This data set is aggregrated by month and joined with the population of each boro to calculate the shooting rate per month. 

The columns of the resulting table are:

* **boro:** Which of the five boros the shooting occurred in.
* **month:** Dates and times of shootings are rounded to the nearest month for aggregation and display purposes. That is, a shooting on (for example) May 13th 2021 is grouped under "May 2021", but one on May 25th 2021 would be rounded up to June 2021. Therefore for display purposes the date of 06/01/2021 encompasses shootings from mid-May to mid-June.
* **population:** The population of the boro (based on the 2020 census).
* **shootings:**  The total number of shootings for the month.
* **rate_per_million:** The shooting rate per million residents, obtained by scaling the total number of shootings in the boro by its population. 

```{r data_cleanup}

# Now group by key to look at distinct events. Create new fields for the number of records
# created for the incident, and whether at least one of the incidents was flagged as a murder
# (that is, someone died as a result of the shooting event)
ny_unique_events <-  ny_initial_processing %>% group_by(key) %>%
    dplyr::summarize(
        date=min(date), 
        boro = first(boro),
        records = n(),
        deaths = sum(flag=="true") >=1,
    )

# Aggregate distinct shootings in each boro monthly, and find the rate per million. Use the last date of 
ny_monthly_by_boro <- ny_unique_events %>% group_by(boro,month = lubridate::round_date(date, "month")) %>%
    summarize(
        shootings=n(),
        boro = first(boro)
    ) %>% left_join(boro_populations,by=join_by(boro)) %>%
    mutate(rate_per_million =shootings/population*1000000)  %>% mutate(boro=as.factor(boro))

ny_totals_per_boro <- ny_monthly_by_boro %>% group_by(boro) %>%
    summarize(
        Total=sum(shootings),
        Boro = first(boro),
        Population = first(population)
    )  

ny_totals_per_boro <- ny_totals_per_boro %>% mutate(Rate=Total/Population*1000000) %>% select(Boro,Total,Population,Rate)



```


## 2.5 Missing Data Handling

Staten island - the smallest boro - had no shootings for eleven months of the time period, and the monthly aggregation steps above did not create rows for these automatically. The months that had zero shootings for Staten Island are added below  - ensuring that each boro has monthly totals for the entire 2006 to 2023 time period.

```{r missing_data }

# Find all the months using one of the other boros.
temp <-  ny_monthly_by_boro %>% filter(boro=="BRONX") %>% select(month)
all_months <- temp$month

# Find the months that Staten Island has a record and use the set difference to get the missing months.

temp <-  ny_monthly_by_boro %>% filter(boro=="STATEN ISLAND") %>% select(month)
staten_months <- temp$month
missing_months <- as.Date(setdiff(all_months,staten_months), origin="1970-01-01")

# Build a data frame with those months.
staten_no_shootings <- data.frame(month=missing_months,boro="STATEN ISLAND", shootings=0,
                                  population=495747,rate_per_million=0)

# Add back in to the monthly data, and rearrange so they are ordered by month.
ny_monthly_by_boro = bind_rows(ny_monthly_by_boro,staten_no_shootings)
ny_monthly_by_boro <- ny_monthly_by_boro  %>% mutate(boro=as.factor(boro))  %>% arrange(boro,month)

```

A summary of the monthly shooting data - to confirm there are no missing monthly totals - is below.

```{r monthly_summary, results="markup"}

summary(ny_monthly_by_boro)
```

 
# 3 Comparing Shooting Incidents Across Boros

This section and the next one explore respectively:

1. Did the rate of shooting incident - adjusted for boro population - differ across the five boros over the time period? 

2. Looking the incident rates in each boro over time, are there any noticeable patterns?

Before looking at plots of monthly shootings, a brief look at overall totals. The table below looks at the total shooting incidents in the five boros from 2006 to 2023. The rate per million residents is also shown. The Bronx has the highest rate, followed by the Brooklyn and then Manhattan. Staten Island and Queens have the lowest rates. The plots in the next section will look at monthly rates over the time period.

```{r totals,echo=FALSE, results="markup" }

knitr::kable(ny_totals_per_boro) %>% kableExtra::kable_styling(position = "center",full_width=F) 

```


The analysis is motivated with visualizations.

## 3.1 Visualizations

A boxplot comparing the monthly shooting incidents per million residents for each boro is below. The mean is also shown as a dot for each boro. Violin plots are also shown, to show the distributions more clearly.


```{r monthly_boxplot}
ggplot(ny_monthly_by_boro, aes(x=boro,y=rate_per_million,fill=boro)) +
    geom_boxplot(color="blue") +
    xlab("Boro") +
    ylab("Monthly Shootings Per million residents") +
    scale_color_brewer(palette="Dark2")+
    stat_summary(fun=mean, geom="point", size=2, color="Blue") +
    theme(legend.position="none") +
    coord_flip()

ggplot(ny_monthly_by_boro, aes(x=boro,y=rate_per_million,fill=boro)) +
    geom_violin(color="blue", trim=FALSE) +
    xlab("Boro") +
    ylab("Monthly Shootings Per million residents") +
    scale_color_brewer(palette="Dark2") +
    stat_summary(fun=mean, geom="point", size=2, color="Blue") +
    theme(legend.position="none") +
    coord_flip()
```

The plots indicate:

* The shooting rates across all boros are not identical. Brooklyn and the Bronx have higher rates thIan the other three.
* Of those three, Manhattan may have a slightly higher rate of shooting incidents, adjusted for population.
* Staten island and Queens appear to have similar incident rates.
* All the boros except Staten Island have right-skewed distributions - the mean is greater than the median. However this apparent skew may be partly due to outliers - all the boros have at least one outlier - in this case, months with an unusually high rate of shooting incidents.
* The Brooklyn and Bronx distributions appear to be more variable than the other three. 

## 3.2 Analysis

To formally test whether the average monthly shooting rates - adjusted for population - are different, a one-way ANOVA (Analysis of Variance) model will be used - see for example [Faraway](https://www.routledge.com/Linear-Models-with-R/Faraway/p/book/9781439887332), chapter 15..
The null hypothesis is that all means are equal. A low p-value provides evidence that at least one mean is different, but does not indicate which one(s).

If the ANOVA indicates at least one mean differs from the others, pairwise tests will be conducted to determine which of those differences are statistically significant.

A brief desciption of the model follows. The "response" $Y$ is the rate of shooting incidents, and the "factor" is which boro the shooting took place in. That is:

$$Y_{ij}=\mu_{j} + \varepsilon_{ij}$$

Where $Y_{ij}$ is the shooting rate for the $i$th month, and $j$ is the index of the boro. The $\varepsilon_{ij}$ are random error terms. The model assumes that these terms are independent, normally distributed, and of equal variance. The next section analyzes these assumptions in detail. In summary, it is found that the model deviates from normality and constant variance. The traditional ANOVA model is quite robust with respect to deviations from normality (see for example [here](https://pubmed.ncbi.nlm.nih.gov/29048317/), but a to adapt for non-equal variances, [Welch's ANOVA](https://www.statology.org/welchs-anova-in-r/) is used.

### 3.2.1 Assumptions of the Model

The ANOVA model used above makes assumptions about the error terms $\varepsilon_{ij}$:

* *Normality*: the error terms (and thus the response) are normally distributed
* *Constant Variance* (or "homoscedasticity"): the error terms have constant variance across observations.
* *Independence*: the error terms are independent (or equivalently, uncorrelated).


As regards *normality*, the boxplot in section 3.1 suggests that the distributions of incident rates over each boro are non-normal. The model is quite robust with respect to this.

The *equal variance* assumption also looks to be violated in the boxplot. Estimating the standard deviation for each boro (below) gives further evidence that this is the case, and the Bartlett test for equal variances shows a statistically significant difference.

The *independence* assumption is harder to test, but given the nature of the data here (shooting incidents in different boros) there is unlikely to be systematic dependence between incidents in different boros (though there may be a few cases of shooting incidents that cross from one boro to another).

In summary, a known issue with the model is that the variance of shootings does not appear to be constant across boros. To adjust for this issue, an ANOVA variant that does not assume equal variances - Welches ANOVA - will be used.

```{r assumptions, results="markup"}

ny_monthly_summary  <- ny_monthly_by_boro %>% group_by(boro) %>% summarize(mean=mean(rate_per_million), std_dev = sd(rate_per_million))


knitr::kable(ny_monthly_summary) %>% kableExtra::kable_styling(position = "center",full_width=F) 

bartlett.test(rate_per_million ~ boro, data=ny_monthly_by_boro )

```
### 3.2.2 Using Welch's One-way ANOVA to Test for Differences in Means

The results of Welch's ANOVA are shown below. A linear model with `rate_per_million` as the reponse and `boro` as factor is constructed, and the model passed to the `oneway.test`function. 

The p value is very close to zero, and thus the null hypothesis of equal means is rejected. This gives evidence that at least one mean is significantly different than the others, but not which one. The next step is to conduct pairwise tests on the differences.

```{r anova,results="markup"}

oneway.test(rate_per_million ~ boro, data = ny_monthly_by_boro, var.equal = FALSE)

```

### 3.2.3 Pairwise Tests of Mean Differences

A potential issue with pairwise tests of popluation means is that since each test has a chance of a type 1 error, conducting multiple tests at once must adjust for this or there will be a large probability of a type 1 error for at least one of the tests. There are a variety of approaches for this, and no consensus as to the best one (this [survey](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7720730/) compares options). We will use one of the most commonly used approaches, the Tukey HSD (Highly Significant Difference) test. This is available in Base R, and (as a bonus) gives confidence intervals for each difference as well as a point estimate. The test is carried out below, using an overall familywise type 1 error of $\alpha=0.05$. 


In order to give context to the differences, the average monthly incident rates for each boro are also shown in a table.

```{r Tukey, results="markup"}

m1 = lm(rate_per_million~boro, data=ny_monthly_by_boro)
a1 = aov(m1)
TukeyHSD(a1)

```
```{r boro_means}


# Show the means themselves, to give context to the differences. 
ny_monthly_summary  <- ny_monthly_by_boro %>% group_by(boro) %>% summarize(mean=mean(rate_per_million))
knitr::kable(ny_monthly_summary) %>% kableExtra::kable_styling(position = "center",full_width=F) 

```

All pairs of differences in means are statistically significant - their adjusted p values are less than 0.05 - with the exception of the difference in shooting rates between Staten Island and Queens. The confidence intervals for the differences in means are also shown, and reinforce the observations from the plots earlier. Brooklyn and the Bronx have substantial differences in their mean incident rates from the other three boros, and of those two, the Bronx has the highest incident rate.

As for the three boros with the lower average incident rates - Staten Island, Queens, and Manhattan: while there are statistically significant differences between Queens and the other two, those differences may not be of practical significance given the large variability from month to month.

How incident rates have varied over time is the focus of the next section.

# 4. Trends in Shooting Rates over Time

Having explored how incident rates vary from boro to boro, we turn to exploring how these incident rates vary over time. The pair of plots below show monthly incident rates - again adjusted for population - of each boro from 2006 to 2023. The second plot is a smoothed version of the first. 

The first plot shows how widely the number of shooting incidents vary from month to month in each boro. The smoothed version suggests that - for all this fluctuation - there was generally a decline in shooting rates across boros from 2010 until about 2018, when they trended upward (from these graphs, the possible exception to this pattern is Staten island).


```{r monthly_trends, warning=FALSE, message=FALSE}



ggplot(ny_monthly_by_boro, aes(x=month,y=rate_per_million,color=boro)) +
    geom_line(se=FALSE,size=.7) +
    xlab("Time") +
    scale_color_brewer(palette="Set2") +
    ylab("Monthly Shootings Per million residents")

ggplot(ny_monthly_by_boro, aes(x=month,y=rate_per_million,color=boro)) +
    geom_smooth(se=FALSE, size=.7) +
    xlab("Time") +
    scale_color_brewer(palette="Set2") +
    ylab("Monthly Shootings Per million residents")

```

Focusing on the most recent 5 years (2019-2023), the smoothed version below shows an increase in shooting rates - across all boros - during the pandemic that peaked in late 2021 and has been trending downward since. 

However the unsmoothed plot shows an additional feature: a dramatic spike in shooting rates in Spring of 2020 and - especially in the Bronx - elevated rates over the following years that now appear to be settling down.

Another noticeable pattern in the time series plots is apparent yearly spikes in the data. An interesting area for further analysis would be to investigate for yearly and/or seasonal cycles in shooting rates.


```{r recent_trends}

ny_recent_monthly_by_boro <- ny_monthly_by_boro %>% filter(month>='2019-01-01')

ggplot(ny_recent_monthly_by_boro, aes(x=month,y=rate_per_million,color=boro)) +
    geom_line(se=FALSE,size=.7) +
    xlab("Time") +
    scale_color_brewer(palette="Set2") +
    ylab("Monthly Shootings Per million residents")

ggplot(ny_recent_monthly_by_boro, aes(x=month,y=rate_per_million,color=boro)) +
    geom_smooth(se=FALSE, size=.7) +
    xlab("Time") +
    scale_color_brewer(palette="Set2") +
    ylab("Monthly Shootings Per million residents")
```

# 5. Potential Sources of Bias

Before turning to the conclusions, some potential sources of bias should be acknowledged.

**Personal Bias:**

* I come from a culture that has very different gun laws (Ireland) and found it very difficult to work on data pertaining to so much gun violence. I found it at times exhausting and draining. I can't imagine how difficult it would be for those who have ever experienced gun violence to engage with this data set.
* I should also acknowledge my ignorance of New York City - its different boros and how the differences in each boro might result in differences in shooting rates. More generally, any recommendations as to how to potentially reduce gun violence not in my area of expertise. 
* However, my lack of prior knowledge of NY boros does at least mitigate any potential biases on studying incident rate across boros - one reason I chose this aspect of the shootings to study.

**Potential sources of Bias in the Data**

* It is extremely likely that there is *nonreporting bias*: that many shooting incidents have gone unreported. For this reason, any tallies of shooting incidents throughout the document are only those that have been responded to by local police and logged as a record.

* There are potentially biases in the racial and demographic data pertaining to the alleged perpetrator of the shooting. Brief descriptions of the race and/or age group of the alleged perpetrator are often given, but there is no information on how this data was collected, or how it was checked for reliability.

* More generally, the raw data was collected by one institution - the NY police departments - and is not cross referenced with any other perspectives in order to build a broader narrative. The records give some extremely limited information about *what* occurred - a shooting took place at this location, at a particular date and time - but no context as to *why* the shooting took place.

# 6. Conclusions

The average rates of monthly shooting incidents per million residents was investigated across the five New York Boros. It was found that there were statistically significant differences over all pairs of boros expect for Staten Island and Queens. The Bronx has the highest shooting rates, followed by Brooklyn, and though there were statistically significant differences between the rates in Manhattan and the other two boros, the high variability of rates from month to month may not make this difference of much practical significance.

Looking at how shooting rates varied over time, it was clear that rates generally decreased from 2010 to the end of that decade, trended upwards for a few years, and then began to decline again. While exploring the causes of this increase is beyond the scope of this report, it is notable that there were spikes in the monthly shooting rates in the first quarter of 2020  - as the Covid-19 pandemic began.  

While any causal conclusions are outside the scope of this report, the analysis does suggest one change to the data source itself that would facilitate its use for incidents with multiple victims. As mentioned in section 2.2, the addition of a secondary key uniquely identifying each row that corresponds to a victim would be advantageous. Further details on this issue are discussed in Appendix A.

# Appendix A: Additional Notes on Incidents with Multiple Victims

For incidents with multiple victims, multiple rows are created with the same incident key. The data dictionary states that "A shooting incident can have multiple victims involved and as a result duplicate INCIDENT_KEY’s are produced. Each INCIDENT_KEY represents a victim but similar duplicate keys are counted as one incident".

However, tallying the number of victims for each incident key and tallying those with the `STATISTICAL_MURDER_FLAG` set to 'TRUE' to count the number of deaths for those incidents (code below) did not produce tallies that agreed with media accounts of the event - found from the date and time of the shooting. Additionally, for many of the incidents the apparent number of deaths were exactly half the total number of victims. Filtering for distinct rows (in case the phrase "similar duplicate rows" meant to only consider unique rows had the same issue. The `STATISTICAL_MURDER_FLAG` appeared to be set to 'TRUE' on half the records. 

I decided that I could not find a way of reliably tallying the mass shooting incidents from the raw data, and instead focused the record on *distinct* shooting incidents. I am curious if others found a way of reliably tallying mass shooting incidents. A secondary key uniquely identifying each victim would make this straightforward.

```{r tallying_multiple_victims}

# Attempt to tally the number of victims, and of those, which ones died. the 'flag' field is the STATISTICAL_MURDER_FLAG which is set to TRUE if the shooting resulted in a death,
# ny_tallies_per_incident <-  ny_initial_processing %>% group_by(key) %>%
#     dplyr::summarize(
#         date=min(date),
#         boro = first(boro),
#         victims = n(),
#         deaths = sum(flag=="true")
#     )
```

# Appendix B: Source of Boro Population Data

For simplicity, a data frame with the  population for each boro was used in the document.

That data frame was created from an excel spreadsheet containing 2020 census data at "https://www.nyc.gov". 

The code to recreate the data frame from the original source is below, and can be uncommented and run if required.

```{r boro_population import}

# This import is commented out. Instead the data is embedded in a data frame at the start of the project. It requires the readxl library.

# import(readxl)

# download excel file and save locally in a /data subdirectory of the current one.
# download.file("https://www.nyc.gov/assets/planning/download/office/planning-level/nyc-population/census2020/nyc_detailed-race-and-ethnicity-data_2020_core-geographies.xlsx", destfile = "./data/boro.xlsx",quiet = TRUE, mode = "wb")

# Read in the excel file.
#boro_data <- read_excel("./data/boro.xlsx")

# Extract the fields pertaining to boro populations, and save them in a data frame.
# boro_populations <- boro_data %>%  filter(...2=="Boro2020") %>%  rename(boro=...3,population=starts_with("2020")) %>% select(boro,population) %>% mutate(boro=toupper(boro))

```

